---
title: "Empirical Research Final Project"
author: "Hassan Fayyaz"
date: "05/24/2022"
output: html_document
---


#Topic: Socioeconomic Factors Affecting Health Insurance In The United States

#Professor Kevin Foster

#Foundations of Empirical Research (ECO B2100)

#The City College of New York

#Introduction:

This research study involves data set of individuals
who have health insurance and those without health insurance. 
Some critical variables of the data set are age, gender, 
marital status, race, education, and household income. 
This research study aims to analyze the impact of these variables
and predict the likelihood of someone having health insurance 
based on the variables. 



```{r}

#Cleaning the environment
rm(list = ls(all = TRUE))

require("ipumsr")

#setting the directory
setwd("C:/Users/hassa/Desktop/Econometrics_Data")

ddi <- read_ipums_ddi("nhis_00002.xml")

data <- read_ipums_micro(ddi)

```

#######################

# Loading Packages

#######################

```{r}
install.packages("tidyverse", dependencies = TRUE)
```

```{r}
install.packages("randomForest")
```

```{r}
install.packages("corrplot")
```

```{r}
install.packages("haven")
```

```{r}
install.packages("gridExtra")
```

```{r}
install.packages("Boruta")
```


```{r}

library(ggplot2)
library(stats)
library(dplyr)
library(randomForest)
library(corrplot)
library(haven)
library(AER)
library(foreign)
library(nnet)
library(reshape2)
library(tidyverse)
library(stargazer)
library(ggeffects)
library(gridExtra)
library(readr)
library(readxl)
library(janitor)
library(skimr)
library(plotly)
library(tidyr)
library(lubridate)
library(rmarkdown)
library(flexdashboard)
library(DT)
library(ggforce)
library(tinytex)

```

#######################

# Viewing Data

#######################

```{r}

str(data)

View(data)

summary(data)

colnames(data)

nrow(data)

```

#######################################################

#Cleaning Data Set

#######################################################

```{r}

data$REGION <- as.factor(data$REGION)
levels(data$REGION) <- c("Northeast","Midwest","South","West")

data$SEX <- as.factor(data$SEX)
levels(data$SEX) <- c("Male","Female","Refused","dont know")

data$SEXORIEN <- as.factor(data$SEXORIEN)
levels(data$SEXORIEN) <- c("NIU","Lesbian or gay","straight","bisexual","something else","dont know","refused","NA")

data$MARST <- as.factor(data$MARST)
levels(data$MARST) <- c("NIU","Married","Married spouse not there","Married spouse NA","Widowed","Divorced","Separated","never married","unknown")

data$RACEA <- as.factor(data$RACEA)
levels(data$RACEA) <- c("white","Black","Aleut Alaskan","American Indian","Asian","Other","refused","not ascertained","unknown")

data$HISPETH <- as.factor(data$HISPETH)
levels(data$HISPETH) <- c("Not Hispanic","Mexican","Other Hispanic","NA")

data$YRSINUS <- as.factor(data$YRSINUS)
levels(data$YRSINUS) <- c("NIU","Less than 1 year in US","1-5 years in US","5-10 years in US","10-15 yr in US","15 or more yr in US","NA")

data$CITIZEN <- as.factor(data$CITIZEN)
levels(data$CITIZEN) <- c("No not US citizen","yes US citizen","refused","NA","dont know")

data$ARMFEV <- as.factor(data$ARMFEV)
levels(data$ARMFEV) <- c("NIU","No never active duty","active only for training","yes ever served in armed forces","refused","NA","dont know")

data$EDUC <- as.factor(data$EDUC)
levels(data$EDUC) <- c("NIU","no school","less than hs","12th grade no diploma","HS diploma","GED","some college","assoc deg in tech or occ","assoc deg academic","bachelors","masters","professional degree","doctoral","refused","dont know")

data$EMPSTAT <- as.factor(data$EMPSTAT)
levels(data$EMPSTAT) <- c("NIU","Employed","not employed","dont know")

data$EMPHI <- as.factor(data$EMPHI)
levels(data$EMPHI) <- c("NIU","no workplace did not offer health insurance","yes workplace offer health insurance","refused","NA","dont know")

data$EMPFT <- as.factor(data$EMPFT)
levels(data$EMPFT) <- c("NIU","parttime","fulltime","refused","NA","dont know")

data$HEALTH <- as.factor(data$HEALTH)
levels(data$HEALTH) <- c("excellent","very good","good","fair","poor","refused","dont know")


is.na(data$HOURSWRK) <- which(data$HOURSWRK > 95)  # hours of work each week
is.na(data$HEIGHT) <- which(data$HEIGHT > 94)  # height in inches
is.na(data$WEIGHT) <- which(data$WEIGHT > 900)  # weight in pounds
is.na(data$BMICALC) <- which(data$BMICALC > 900)  # BMI Body Mass Index

data$HINOTCOVE <- as.factor(data$HINOTCOVE)
levels(data$HINOTCOVE) <- c("has health insurance coverage","no health insurance coverage","dont know")

data$EMPSTAT<-as.factor(data$EMPSTAT)
levels(data$EMPSTAT)<-c("not employed","employed","dont know","NIU")

```

#Data cleaning
```{r}
library(dplyr)

mydata= select(data,-c("YEAR","SERIAL","STRATA","PSU","NHISHID","NHISPID","HHX"))
```

#Inspect the data and check for missing values
```{r}
#NA count of all
table(is.na(mydata))

#NA count Column wise
sapply(mydata, function(x) sum(is.na(mydata)))

#NA count Row wise
rowSums(is.na(mydata))

sum(is.na(mydata))
mydata= na.omit(mydata)#remove all missing values
```

######################################

 'Exploratory Data Analysis'

######################################

If the calculated value is greater than the critical value, 
then we need to reject the null hypothesis & if the computed 
value is less than the critical value, then we fail to reject 
the null hypothesis


```{r}

library(sjPlot)

```

######################################
#1 Health Insurance By Race
######################################

```{r Table#1}
sjPlot::tab_xtab(var.row=mydata$RACEA, 
                 var.col =mydata$HINOTCOVE,
                 title = "Health Insurance By Race",
                 show.row.prc = TRUE)
```

```{r Chart#1}
ggplot(mydata,
      aes(x= RACEA, y=HINOTCOVE)) +
  geom_bar(stat = "identity", width=0.5, color="Green") + labs(x = "Race", y = "Health Insurance Status")+
  geom_line(color = "firebrick", linetype = "dotted", size = .3) +
  ggtitle("Health Insurance By Race")
```

######################################
#2 Health Insurance By Sex
######################################

```{r Table#2}

sjPlot::tab_xtab(var.row=mydata$SEX, 
                 var.col =mydata$HINOTCOVE,
                 title = "Health Insurance By Sex",
                 show.row.prc = TRUE,drop.empty = TRUE)
```

```{r Chart#2}
sjPlot::plot_xtab(mydata$HINOTCOVE, mydata$SEX, type = c("bar", "line"),
                  margin = "row",  title = "Health Insurance By Sex",
                  bar.pos = "stack", coord.flip = TRUE)

```

#######################################
#3 Health Insurance By Citizenship Status
#######################################

```{r Table#3}

sjPlot::tab_xtab(var.row = mydata$CITIZEN, 
                 var.col = mydata$HINOTCOVE,
                 title = "Health Insurance By Citizenship Status",
                 show.row.prc = TRUE)
```

```{r Chart#3}
ggplot(mydata,
      aes(x= CITIZEN, y=HINOTCOVE)) +
  geom_point(color = "firebrick", shape = "diamond", size = 2)+ labs(x = "Citizen", y = "Health Insurance Status")+
  geom_line(color = "firebrick", linetype = "dotted", size = .3) +
  ggtitle("Health Insurance by Citzenship")
```

########################################
#4 Health Insurance By Sexual Orientation
########################################

```{r Table#4}

sjPlot::tab_xtab(var.row = mydata$SEXORIEN, 
                 var.col = mydata$HINOTCOVE,
                 title = "Table Health insuracne by sex orientation",
                 show.row.prc = TRUE)
```

```{r Chart#4}
ggplot(mydata, aes(x=SEXORIEN, y=HINOTCOVE)) + 
  geom_bar(stat = "identity", width=0.5, color="Purple") + labs(x = "Sexual Orientation", y = "Health Insurance Status")+
  geom_line(color = "firebrick", linetype = "dotted", size = .3) +
  ggtitle("Health Insurance By Sexual Orientation")
```

########################################
#5 Health Insurance Based On US Military Service
########################################

```{r Table#5}
sjPlot::tab_xtab(var.row = mydata$ARMFEV, 
                 var.col = mydata$HINOTCOVE,
                 title = "Health Insurance Based On US Military Service",
                 show.row.prc = TRUE)
```

```{r Chart#5}
ggplot(mydata, aes(x=ARMFEV, y=HINOTCOVE)) + 
  geom_bar(stat = "identity", width=0.5, color="Pink") + labs(x = "Military Service", y = "Health Insurance Status")+
  geom_line(linetype = "dotted", size = .3) +
  ggtitle("Health Insurance Based On US Military Service")
```

###############################################################
#6 Health Insurance By Years Lived In The United States
###############################################################

```{r Table#6}
sjPlot::tab_xtab(var.row = mydata$YRSINUS, 
                 var.col = mydata$HINOTCOVE,
                 title = "Health Insurance By Years Lived In The United States",
                 show.row.prc = TRUE)
```

```{r Chart#6}
ggplot(mydata, aes(fill=HINOTCOVE, y=YRSINUS, x=HINOTCOVE)) + 
  geom_bar(position="dodge", stat="identity") + labs(x = "Health Insurance Status", y = "Years Lived In The United States") +
  ggtitle("Health insurance by RACE")
```

##########################################################
#7 Health Insurance By Employment Status
##########################################################

```{r Table#7}

sjPlot::tab_xtab(var.row = mydata$EMPSTAT, 
                 var.col = mydata$HINOTCOVE,
                 title = "Health Insurance By Employment Status",
                 show.row.prc = TRUE)
```

```{r Chart#7}
ggplot(aes(x= EMPSTAT, 
           y= HINOTCOVE, 
           color=EMPSTAT), data = mydata)+
  geom_point()+
  geom_smooth(method="lm", se=FALSE) + 
  facet_wrap(~EMPSTAT + HINOTCOVE) +
  ggtitle("Health Insurance By Employment Status")
```

############################################
#8 Health Insurance By US Born Citizenship Status 
############################################

```{r Table#8}

sjPlot::tab_xtab(var.row = mydata$USBORN, 
                 var.col = mydata$HINOTCOVE,
                 title = "Health Insurance By US Born Citizenship Status",
                 show.row.prc = TRUE)
```

```{r Chart#8}
ggplot(mydata,
       aes(x= USBORN, y =HINOTCOVE)) +
  geom_point(color = "firebrick", shape = "diamond", size = 2)+
  ggtitle("Health Insurance By US Born Citizenship Status")+
  xlab("USBORN")+
  ylab("HINOTCOVE")
```

#############################################
#9 Health Insurance By Whether Workplace Offers Health Insurance 
#############################################

```{r Table#9}

sjPlot::tab_xtab(var.row = mydata$EMPHI, 
                 var.col = mydata$HINOTCOVE,
                 title = "Health Insurance By Whether Workplace Offers Health Insurance",
                 show.row.prc = TRUE)
```

```{r Chart#9}
ggplot(aes(x= EMPHI, y=HINOTCOVE), data=mydata)+ geom_point()+
  facet_grid(HINOTCOVE ~ EMPHI) +
  xlab("Workplace Offers Health Insurance") + 
  ylab("Health Insurance Status")

```

##################################################
#10 Health Insurance By Health Status 
##################################################

```{r Table#10}

sjPlot::tab_xtab(var.row = mydata$HEALTH, 
                 var.col = mydata$HINOTCOVE,
                 title = "Health Insurance By Health Status",
                 show.row.prc = TRUE)
```

```{r Chart#10}
ggplot(mydata, aes(x=HEALTH, y=HINOTCOVE)) + 
  geom_bar(stat = "identity", width=0.2, color="Black")+
  xlab("Health Condition") + 
  ylab("Health Insurance Status")+
  ggtitle("Health Insurance By Health Status")
```

################################################################
#11 Health Insurance By Medicaid Status
################################################################

```{r Table#11}
sjPlot::tab_xtab(var.row = mydata$HIMCAIDE, 
                 var.col = mydata$HINOTCOVE,
                 title = "Health Insurance By Medicaid Status",
                 show.row.prc = TRUE)
```

```{r Chart#11}
ggplot(mydata,
       aes(x= HIMCAIDE, y= HINOTCOVE)) + geom_point(color= "Black", alpha = 2) + labs(x = "Medicaid Status", y = "Health Insurance Status") + 
  
stat_smooth() +
theme(panel.grid.major = element_line(size = .1, linetype = "dashed"),
        panel.grid.minor = element_line(size = .2, linetype = "dotted"),
        panel.grid.major.x = element_line(color = "red1"),
        panel.grid.major.y = element_line(color = "blue1"),
        panel.grid.minor.x = element_line(color = "red4"),
        panel.grid.minor.y = element_line(color = "blue4"))+
  ggtitle("Health Insurance By Medicaid Status") 
  
  xlab("")
```

############################################
#12 Health Insurance By Medicare Status
############################################

```{r Table#12}

sjPlot::tab_xtab(var.row = mydata$HIMCAREE, 
                 var.col = mydata$HINOTCOVE,
                 title = "Health Insurance By Medicare Status",
                 show.row.prc = TRUE)
```

```{r Chart#12}
ggplot(mydata,
       aes(x= HIMCAREE, y= HINOTCOVE)) + geom_point() + labs(x = "Medicare Status", y = "Health Insurance Status")+
  geom_violin(draw_quantiles   = c(.25, .5, .75))+
  geom_jitter(aes(y= HINOTCOVE,
                  x= HIMCAREE),
              color="Purple",
              height = 0,
              width=0.5, alpha=100)+
  
ggtitle("Health Insurance By Medicare Status")
xlab("")
```

###############################################
#13 Health Insurance By Years Lived In The United States
###############################################

```{r Table#13}

sjPlot::tab_xtab(var.row = mydata$YRSINUS, 
                 var.col = mydata$HINOTCOVE,
                 title = "Health Insurance By Years Lived In The United States",
                 show.row.prc = TRUE)
```

```{r Chart#13}
ggplot(aes(x= YRSINUS, 
           y= HINOTCOVE, 
           color=YRSINUS), data = mydata)+
  geom_point()+
  geom_smooth(method="lm", se=FALSE) + 
  facet_wrap(~YRSINUS + HINOTCOVE) +
  ggtitle("Health Insurance by Years lived in the United States")
```

#######################################

# LOGISTIC REGRESSION

#######################################

# Logistic Regression

# The dependent variable health insurance "HINOTCOVE" has 3 levels, thus will be recoded 
# to include 2 levels in order to fit a logistic regression model


```{r}
mydata$H_insurance = ifelse(mydata$HINOTCOVE == "has health insurance coverage","has  insurance coverage","no insurance") #drop the - HINOTCOVE
colnames(mydata)
```
#Lets inspect our variable and change it to a factor

```{r}

table(mydata$H_insurance)
class(mydata$H_insurance)
mydata$H_insurance = as.factor(mydata$H_insurance)
```

# Drop irrelevant variables
```{r}
mydata = mydata[,-c(1:9)]

dim(mydata) #this is a very big dataset with 31176 rows and 45 columns
```
# NOTE: We will use a small sample for the modelling to utilize the small PC RAM 
# Sample 10% of the data

```{r}
sample_size = floor(0.01*nrow(mydata))
set.seed(777)
picked = sample(seq_len(nrow(mydata)),size = sample_size)
mydata_sample =mydata[picked,]
dim(mydata_sample)
```

# Feature selection using BORUTA PACKAGE
```{r}
library(Boruta)
set.seed(123)
boruta.train <- Boruta(H_insurance~. ,data = mydata_sample, doTrace = 2)
print(boruta.train)
```

```{r}
plot(boruta.train, xlab = "", xaxt = "n")
lz<-lapply(1:ncol(boruta.train$ImpHistory),function(i)
    boruta.train$ImpHistory[is.finite(boruta.train$ImpHistory[,i]),i])
names(lz) <- colnames(boruta.train$ImpHistory)
Labels <- sort(sapply(lz,median))
axis(side = 1,las=2,labels = names(Labels),
     at = 1:ncol(boruta.train$ImpHistory), cex.axis = 0.7)
```

```{r}
final.boruta <- TentativeRoughFix(boruta.train)
print(final.boruta)
```

```{r}
boruta.df <- attStats(final.boruta)
head(boruta.df)

features=getSelectedAttributes(final.boruta, withTentative = F)
features
```

```{r}
final.features=c("bhr","basedp","pkhr","X.mphr.b.","sbp","dp","maxhr","mbp",
                 "dpmaxdo","age","gddpeakdp")
my.features=mydata_sample[,features]
```




# Re-introduce the y class variable

```{r}

my.features$H_insurance = mydata_sample$H_insurance

```

###########################################

 Now lets fit a logistic regression model
 
###########################################

```{r}

mymodel =glm(H_insurance ~ .,data=my.features,family = "binomial")
summary(mymodel)
```



# Now lets fit a logistic regression model

```{r}

mymodel = glm(H_insurance ~.
     ,data=my.features,family = "binomial")

summary(mymodel)

```

# Odds ratios
```{r}
exp(mymodel$coefficients[c(3,29,69,70,72:76)])
```

# Print the model

```{r}
print(mymodel)
```
# Odds ratios


```{r}
exp(mymodel$coefficients[c(3,29,69,70,72:76)])
# Print the model
tab_model(mymodel)

```
#########################################

# Multinomial Logistic Regression

#########################################

```{r}
install.packages("packagename")
```

```{r}
require(foreign)
require(nnet)
require(ggplot2)
require(reshape2)
```

```{r}
data$HINOTCOVE <- relevel(factor(data$HINOTCOVE), ref = "has health insurance coverage")
test <- multinom(data$HINOTCOVE ~ data$RACEA + data$YRSINUS +
                   data$SEX + data$MARST + data$EDUC)

```

```{r}
exp(mymodel$coefficients[c(3,29,69,70,72:76)])
```

```{r}
model = summary(test)
z <- summary(test)$coefficients/summary(test)$standard.errors
z
```

```{r}
exp(coef(test))
```
```{r}
head(pp <- fitted(test))
```

```{r}
dses <- data.frame(insurance = c("has health insurance coverage","Employed"),
                   write = mean(mydata$write))

```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```


```{r cars}
summary(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
